Hippocampal replay of experience at real-world speeds

### 2020-10-05 Article.tex
Update on Overleaf.

\title{Modeling the movement dynamics of hippocampal replay reveals slower spatially coherent trajectories}
\shorttitle{Replay Dynamics}

@@

%TC:break Abstract
%the command above serves to have a word count for the abstract
\begin{abstract}
Representations of past and possible future experiences play a critical role in memory and decision-making processes. The hippocampus is known to express such representations during sharp-wave ripple (SWR) events, when ensembles of hippocampal place cells can be activated in time-compressed sequences consistent with prior experiences. Efforts to understand these "replay" events have typically imposed stringent criteria on SWR event significance, and thus analyze only a small subset of events (\texttildelow10-45\%). As a result, we lack a clear understanding of both the prevalence and the range of representational dynamics associated with replay. Here, we develop a state space model that uses a combination of movement dynamics of different speeds to capture the spatial content and time evolution of replay during SWR events. Application of this model reveals that 85\% of events contained spatially coherent, interpretable content. Furthermore, over 60\% of SWR events contain slower spatially coherent trajectories or representations of single locations, challenging the notion that SWRs contain primarily replay of high speed trajectories.

@@

The brain has the remarkable ability to retrieve representations of past events and generate representations of possible future events. These internal "generative" representations \cite{KayConstantSubsecondCycling2020} can be identified because the underlying neural activity patterns resemble those seen during actual events. In particular, an encoding model that relates neural firing to external covariates can be constructed during an experience, and this model can be inverted to decode an internal representation of the covariates from the neural activity at a different time.

Mental events may not occur in the same time span as actual events: memories can be retrieved and used in less time than was required for the original experience, and mental simulations can span a series of actions more quickly than would be required to perform those actions. Consistent with this, generative representations can unfold rapidly \cite{NadasdyReplayTimeCompression1999, LeeMemorySequentialExperience2002, DavidsonHippocampalReplayExtended2009, KarlssonAwakereplayremote2009}. As a result, decoding generative representations often involves making assumptions about their timescale and temporal structure, and these assumptions can limit the interpretation of the data.

Hippocampal replay events are a prototypical example of a generative representation. As animals move through space, neurons in the hippocampus preferentially fire at specific locations in an environment, and thus sets of cells fire in sequence as the animal moves through a series of locations. When the animal is asleep or immobile, hippocampal cells can be reactivated during a "sharp-wave ripple" (SWR) event. A subset of SWRs contain sequential firing similar to that seen during a previous experience or likely to occur during potential future experiences, and previous work has reported that these sequential firing events proceed at an average speed of ~10 meters per second, about 20x faster than the animal's usual movement speed \cite{NadasdyReplayTimeCompression1999, LeeMemorySequentialExperience2002, DavidsonHippocampalReplayExtended2009, KarlssonAwakereplayremote2009}.

Identifying these events typically involves multiple steps and assumptions about the nature of replay, which are mostly commonly characterized using the standard "Bayesian" decoder. First, an encoding model is constructed based on spiking activity during movement, most often using only spikes that have been clustered as single units (putative single neurons). Then, a subset of SWRs or events with high activity levels are selected based on a threshold for event size chosen by the experimenter \cite{FosterReversereplaybehavioural2006, DibaForwardreversehippocampal2007, KarlssonAwakereplayremote2009, StellaHippocampalReactivationRandom2019}. The spikes within these events decode the probability of position in overlapping or non-overlapping time bins whose size is also chosen by the experimenter. Finally, the most commonly used approach to detecting sequential replay involves fitting a line to the resulting set of probability distributions, which relies on the assumption that the representations smoothly progress at a constant speed \cite{FosterReversereplaybehavioural2006, DibaForwardreversehippocampal2007, KarlssonAwakereplayremote2009}. A statistical test is then used to determine whether the fit is better than the fit to shuffled versions of the data.

While the standard decoder approach identifies constant speed events, it does not consider events that are rejected by the statistical test, and the use of a fixed size temporal bin acts as a boxcar smoother that limits the potential movement speeds of the representation. The linear fit is also problematic, because it has the potential to reject real events that do not move at constant speeds. Approaches that focus on the order of cell activity within each event \cite{LeeMemorySequentialExperience2002, GuptaHippocampalReplayNot2010} allow for a relaxation of that linear assumption, but replace it with a loss of statistical power due to either ignoring all but the first spike from each cell or using an arbitrarily smoothed version of the spike train. These approaches also do not provide information about the dynamics of the underlying spatial representation. Moreover, there is now evidence that a subset of these events have stationary representations of a single location \cite{JaiDistincthippocampalcorticalmemory2017, FarooqEmergencepreconfiguredplastic2019}, which both approaches often exclude. In addition, these methods identify only a subset of events as containing content (typically somewhere between 10 and 45\% of the chosen events), and thereby provide little insight into the remaining majority of the events.

Recognizing the problems with the linear fit, several studies have moved away from the constant velocity assumption, using an estimated represented location at each time bin and connecting each with a line. For example, using this approach, Pfeiffer and Foster \cite{PfeifferAutoassociativedynamicsgeneration2015} found that replays can alternate between representing a single location and sequential spatial trajectories. On the other hand, Stella et al. \cite{StellaHippocampalReactivationRandom2019} found that replays are spatially continuous, following Brownian diffusion dynamics. Both methods still used large time bins and neither took into account the uncertainty of the decoded estimates, however, making it hard to identify the source of the different conclusions.

An ideal approach to identifying and quantifying the dynamics of generative representations would circumvent these problems. It would use all of the available spiking data to yield the most accurate decoded positions. It would be applicable to either thresholded events or to all of the data to permit an unbiased assessment of the prevalence and nature of generative activity. It would use very small temporal bins (1 or 2 ms) to allow for very rapid representational movement and provide information about the certainty of the decoded estimates. It would be able to capture different types of movement dynamics, ranging from stationary to rapidly evolving to disorganized, and provide a statistical assessment of confidence for each category of dynamics. Finally, where assumptions are made, it would provide well defined parameters whose values can be explored systematically to understand their influence on the results.

We therefore developed a state space model of generative activity that achieves all of those goals. State space models are a well-understood, well-known statistical solution to the problems described above. By mathematically modeling the relationship between the data and latent dynamics, state space models make the assumptions of the model explicit and interpretable. Our model goes beyond previous approaches \cite{MaboudiUncoveringtemporalstructure2018, DengRapidclassificationhippocampal2016} by characterizing represented trajectories as a mixture of three underlying patterns of movement dynamics: stationary trajectories, continuous trajectories that are many times the typical speed of the animal, and spatially fragmented trajectories. We show how this model can take advantage of clusterless decoding---which relates multiunit spike waveform features to position without spike sorting---giving us more information about the population spiking activity. We apply this model to data from 10 rats, focusing on SWR events to permit a direct comparison to previous work.

We find that the majority of SWRs contain spatially coherent content; that is, the replay trajectories have no large jumps. Surprisingly, while the expected high speed replay events were identified, the most common category of events expressed representations that moved at much slower speeds, more consistent with stationary or slower moving trajectories, suggesting that high speed replay is not the most common manifestation of generative activity. These findings illustrate the power of state space models and provide a new understanding of the nature of hippocampal replay.

@@

Next, we specify movement dynamic models for how latent position---the "mental" position of the animal represented by the cells---evolves over time for a given dynamic (Figure \ref{1}C). We do this by defining a state transition matrix, which defines how the latent position can move from the previous time step. Previous findings suggest that replay may exhibit at least three distinct types of movement dynamics: stationary \cite{JaiDistincthippocampalcorticalmemory2017, FarooqEmergencepreconfiguredplastic2019}, continuous \cite{DavidsonHippocampalReplayExtended2009}, and fragmented, which could correspond to both extended spatially incoherent representations and representations that jump from one position to another in a short time period \cite{PfeifferAutoassociativedynamicsgeneration2015}. We therefore define state transition models to capture each of these movement dynamics.

@@

Because the model is defined in terms of a mixture of movement dynamics, as summarized by the probability of each movement dynamic, this allows the model to capture a wide range of movement speeds for the latent position. To demonstrate this, we applied the model to 10,000 simulated trajectories, each trajectory proceeding at a constant speed (Figure \ref{1}F) from 1 cm/s to 10,000 m/s. From this, we can see that not only are there regions of speed that correspond to our three movement dynamics being highly probable (where we define highly probable to be greater than or equal to 0.8 probability), there are intermediate speeds where two of the dynamics exhibit relatively high probability; and where the sum of two of the dynamics' probabilities exceeds 0.8. In this manuscript, we will refer to these as mixture dynamics. For example, when the stationary dynamic has a probability of 0.6 and the continuous has a probability of 0.4, we call this a stationary-continuous mixture (light blue, Figure \ref{1}G) and this indicates that the trajectory is moving slowly. Correspondingly, if the continuous dynamic has a probability of 0.5 and the fragmented dynamic has a probability of 0.4, then we would call this a fragmented-continuous-mixture and this indicates the trajectory is moving very quickly, but not as quickly as the fragmented dynamic dictates. By using this classification scheme, we can characterize the speed or set of speeds in a replay by its movement dynamics. To summarize, there are three movement dynamics, each of which is assigned a probability over time. We further classify the probability of each movement dynamic as being part of one of five speed categories: stationary, stationary-continuous-mixtures, continuous, fragmented-continuous mixtures, and fragmented.

@@

\begin{figure*}%[tbhp]
\centering
\includegraphics[width=0.80\linewidth]{figures/Figure2/Figure2_v5}
\caption{The model can decode hippocampal replay trajectories using either sorted and clusterless spikes from the same SWR event. \textbf{(A)} Decoding using sorted spikes. The top panel shows 31 cells on a W-track ordered according to linearized position by their place field peaks. The second panel shows the probability of each dynamic over time as in Figure \ref{1}F. Shaded regions correspond to the speed classifications as in Figure \ref{1}G. The third panel shows the estimated probability of latent position over the course of the SWR as it travels down the center arm toward the right arm. L, R, C correspond to the position of the left, right and center reward wells respectively. The animal's actual position is indicated by the the magenta dashed line. Underneath is the maximum of the 1D decoded position (the most probable position) projected back onto the 2D track for the sorted decoding. Color indicates time. The animal's actual position is denoted by the pink dot. Light grey lines show the animal's 2D position over the entire recording session. \textbf{(B)} Decoding using clusterless spikes. The top panel shows multiunit spiking activity from each tetrode. Other panels have the same convention as (A). Underneath is the maximum of the 1D decoded position (the most probable position) projected back into 2D using the clusterless decoding.
}
\label{2}
\end{figure*}
\subsection*{Identification of continuous replays with sorted and clusterless spikes}

\begin{figure*}%[tbhp]
\centering
\includegraphics[width=0.80\linewidth]{figures/Figure3/Figure3_v5}
\caption{A-F. Six examples of SWRs with non-constant speed trajectories. Figure conventions are the same as in Figure \ref{2}. Filtered SWR ripple (150-250 Hz) trace from the tetrode with the maximum amplitude displayed above each example. \textbf{(A)} A SWR where the decoded position starts moving down the center arm away from the animal's position at the center well, slows down, and returns back. \textbf{(B)} A SWR where the decoded position persistently stays at the choice point (open circle) while the animal remains at the left well. \textbf{(C)} A SWR where the decoded position begins with stationary representation of the left well, then jumps to the middle of the right arm and proceeds up the right arm to the right well. \textbf{(D)} A SWR where the decoded position begins with stationary representation of the left well, jumps to the center arm, proceeds away from the center well, jumps to the right arm, proceeds back toward the center well, and then becomes fragmented. \textbf{(E)} A SWR where the decoded position begins in the left arm and persists at the choice point. \textbf{(F)} A SWR where the decoded position starts in the left arm toward the choice point, jumps to the right arm and proceeds back toward the choice point. \textbf{(G)} Each dot represents the percentage of SWRs for each day for that animal. A SWR is included in the numerator of the percentage if at any time it includes the classifications listed below the column. The denominator is listed in the x-axis label.
}
\label{3}
\end{figure*}

@@

\subsection*{Replays with non-constant speed}

After testing our model on a single SWR event, we applied our decoding algorithm to hippocampal recordings from 10 animals performing the W-track spatial alternation task (\# tetrodes range: [10, 24], brain areas = [CA1, CA2, CA3]; some data previously used in \cite{KarlssonAwakereplayremote2009, KayConstantSubsecondCycling2020, CarrTransientSlowGamma2012}). One major goal of our work was to assess the overall prevalence of spatial content across SWRs, so we detected SWRs using a relatively permissive threshold (see Methods) to ensure that we included both the rare large events as well as the much more common small events.

@@

Our model is specified in the context of a latent position associated with different movement dynamics. This specification allows us to not only to classify events in terms of their dynamics, but also to quantify the model's certainty in each position estimate at each moment in time given the model parameters. To do so we can compute the cumulative spatial extent of the 95\% highest posterior density (HPD) region of the posterior linear position. Larger values of the cumulative HPD extent indicate the model is less certain about position, because the most probable decoded positions are distributed over more of the track at a given time point. In contrast, lower values indicate that the model is more certain about the estimate of position because the extent of the HPD is more concentrated and covers less of the track. Thus, the HPD provides a complementary measure of spatial coherence, and evaluating it allows us to verify that the events we defined as spatially coherent correspond to events where there is high certainty around the position estimates. Although this is certainly influenced by the dynamics imposed on the model via the movement state transition matrix, because the model is primarily driven by the data, it does not necessarily have to be the case that the two coincide. To show this, as before, we resampled position with replacement for two recording sessions, shuffling the relationship between spiking and spatial information and then fitting the encoding model. The shuffled decodes had much larger spatial coverage than the real data (XX cm vs XX cm, p=2.2e-16, one-sided Mann-Whitney-U, \ref{fig:Figure4-Figure supplement 1}).

We found that spatially coherent events did indeed have lower cumulative HPD sizes than events with fragmented dynamics. Figure \ref{4}A and B show two example SWRs that were classified as having stationary and continuous dynamics, respectively. The bulk of the HPD values at each time step in these SWRs is concentrated in a small portion of the track and the cumulative HPD size is low throughout the SWRs. In contrast, Figure \ref{4}C shows a SWR where the dynamics are fragmented and correspondingly, the HPD is much more spatially diffuse and the cumulative HPD size is much higher. The HPD also provides insights into the moment-by-moment structure of each event, which can change over the time course of a SWR. An example of this change is shown in Figure \ref{4}D, where the cumulative HPD size is small for most of the SWR until the end, at which point the uncertainty of position becomes much higher, reflecting a transition from a spatially coherent to a spatially incoherent representation. Overall, when we examined the average cumulative HPD size for each SWR, grouped by dynamic, we found a clear bimodal split between spatially coherent dynamics and spatially incoherent dynamics (Figure \ref{4}E,). For the spatially coherent dynamics, the cumulative HPD was much lower than the spatially incoherent dynamics (median 24 cm, spatially coherent vs. median 238 cm, spatially incoherent, p=2.2e-16, one-sided Mann-Whitney-U).

\begin{figure*}%[tb2]
\centering
\includegraphics[width=0.80\linewidth]{figures/Figure4/Figure4_v6}
\caption{Validation of the Dynamics using the 95\% Highest Posterior Density. \textbf{(A-D)}  Examples of the 95 \% Highest Posterior Density. Top panel: Probability of dynamic over time. Shading and labels indicate dynamic categories. 2nd panel: Posterior distribution of estimated linear position over time. Red triangle indicates animal's position. 3rd panel: Posterior distribution plot, where black designated the position bins that are included in the 95\% highest posterior density (HPD) region. 4th panel: Cumulative HPD size. \textbf{(E)} Average Cumulative 95\% HPD size for each dynamic category.}
\label{4}
\end{figure*}

\subsection*{Many replay trajectories have slower movement speeds}
\begin{figure*}%[tbhp]
\centering
\includegraphics[width=0.80\linewidth]{figures/Figure5/Figure5_v4}
\caption{
Quantification of the SWRs by dynamic. \textbf{(A)} UpSet plot \cite{LexUpSetVisualizationIntersecting2014}---which is similar to a Venn diagram with more than 3 sets---of the most common sets of classifications within each SWR. Each row represents a classification and each column represents a set of classifications, where filled-in black dots with an edge between the dots indicates that multiple classifications are present in the SWR (at temporally distinct times). The sets are ordered by how often they occur as indicated by the bar plot above each category. The total percentage of each classification is indicated by the rotated bar plot on the left. \textbf{(B)} Box plot of the duration of each dynamic category within a SWR. The box shows the interquartile range (25-75\%) of the data and the whiskers are 1.5 times the interquartile range. Points outside of this are labeled as diamonds. \textbf{(C)} Box plot of the average distance between the latent position to the animal's position for each classification within the SWR. \textbf{(D)} Box plot of the average speed of the classification within the SWR (excluding classifications < 20 ms).\textbf{(E)} Kernel density estimate of the position of stationary trajectories on the W-track at least 30 cm away from the animal's position. The shaded region represents the density estimate while the black ticks represent the observed non-local stationary positions. \textbf{(F)} Box plot of the average tetrode multiunit spike rates for each dynamic category within each SWR (excluding classifications < 20 ms).
}
\label{5}
\end{figure*}

Surprisingly, our results indicate that most events with coherent spatial content are not dominated by the continuous movement dynamic, and thus correspond to trajectories that are stationary or that move relatively slowly compared to their purely continuous counterparts. We therefore examined these events in more detail. We first note that most of the SWRs (16454 of 23382 classified SWRs or 70\%) are well described by a single dynamic. Surprisingly, the most common category is stationary-continuous mixtures (9860 of 23382 classified SWRs or 42\%, Figure \ref{5}A). These events contain representations that move at rates more consistent with real-world movement (Figure \ref{1}G), and are slightly shorter, on average, than events with continuous dynamics (median duration: stationary-continuous-mixture 72 ms vs. continuous 95 ms, Figure \ref{5}B). Nonetheless, both these slow-moving events and continuous events represented locations that were some distance away from the animal (mean trajectory distance from animal's position: stationary-continuous-mixture 51 cm vs. continuous 42 cm, Figure \ref{5}C). This indicates that the trajectories of these SWRs, like ones that have faster continuous dynamics, do not represent the animal's actual position and cannot be explained by erroneously detected SWRs.

>>>>

\title{Hippocampal replay of experience at real-world speeds}
\shorttitle{Replay Dynamics}

@@

%TC:break Abstract
%the command above serves to have a word count for the abstract
\begin{abstract}
Representations of past and possible future experiences play a critical role in memory and decision-making processes. The hippocampus expresses these sorts of representations during sharp-wave ripple (SWR) events, and previous work identified a minority of SWRs that contain “replay” of spatial trajectories at ~20x real-world speeds. Efforts to understand replay typically make multiple assumptions about which events to examine and what sorts of representations constitute replay, however. We therefore lack a clear understanding of both the prevalence and the range of representational dynamics associated with replay. Here we develop a state space model that uses a combination of movement dynamics of different speeds to capture the spatial content and time evolution of replay during SWRs. We find that the large majority of replay events contain spatially coherent, interpretable content. Furthermore, most events progress at real-world, rather than accelerated, movement speeds, consistent with actual experiences.

@@

The brain has the remarkable ability to retrieve representations of past events and generate representations of possible future events. These "generative" \cite{KayConstantSubsecondCycling2020} mental events may not occur in the same time span as actual events: memories can be retrieved and used in less time than was required for the original experience, and mental simulations can span a series of actions more quickly than would be required to perform those actions. Thus, a potential neural substrate for retrieval and mental simulation would be expected to support rapid reinstatement of patterns of brain activity associated with past or potential future experience

Hippocampal "replay" events are well suited to that role \cite{CarrHippocampalreplayawake2011}. As animals move through space, neurons in the hippocampus preferentially fire at specific locations in an environment, and thus sets of cells fire in sequence as the animal moves through a series of locations. When the animal is asleep or immobile, hippocampal cells can be reactivated during a "sharp-wave ripple" (SWR) event. A subset of SWRs contain sequential firing similar to that seen during a previous experience, and are thus thought to "replay" these previous experiences. Importantly, previous work has reported that these sequential firing events proceed at an average speed of ~10 meters per second, about 20x faster than the animal's usual movement speed \cite{NadasdyReplayTimeCompression1999, LeeMemorySequentialExperience2002, DavidsonHippocampalReplayExtended2009, KarlssonAwakereplayremote2009}.

While the existence of these sequential events is well established, the current consensus is that only a minority (perhaps ~10-45\%) of hippocampal SWRs contain statistically identifiable, sequential replay. Of the remaining events, some contain spiking of neurons associate with single locations where animals are immobile (Cite Yu et. al.), but another report suggested that events that replay a single location are only seen in young animals (Cite Stella). Thus, while it would seem useful to replay memories associated with single locations and perhaps even experiences occurring at real-world speeds, the prevalence of that sort of replay in awake animals remains unclear.

Our uncertainty stems in part from the dominant approaches used to identify the content of replay events. These approaches typically involve multiple steps and multiple assumptions about the nature of replay, which are mostly commonly characterized using the standard "Bayesian" decoder. First, an encoding model is constructed based on spiking activity during movement, most often using only spikes that have been clustered as single units (putative single neurons). Then, a subset of SWRs or events with high activity levels are selected based on a threshold for event size chosen by the experimenter \cite{FosterReversereplaybehavioural2006, DibaForwardreversehippocampal2007, KarlssonAwakereplayremote2009, StellaHippocampalReactivationRandom2019}. A decoding algorithm is then applied to the  spikes within these events, yielding a set of probability distributions. Current approaches use either overlapping or non-overlapping time bins whose size is also chosen by the experimenter. Finally, the most commonly used approach to detecting sequential replay involves fitting a line to the resulting set of probability distributions, which relies on the assumption that the representations progress at a constant, non-zero speed \cite{FosterReversereplaybehavioural2006, DibaForwardreversehippocampal2007, KarlssonAwakereplayremote2009}. A statistical test is then used to determine whether the fit is better than the fit to shuffled versions of the data.

While the standard decoder approach identifies constant speed events, it does not consider events that are rejected by the statistical test, and the use of a fixed size temporal bin acts as a boxcar smoother that limits the potential movement speeds of the representation. The linear fit is also problematic: it has the potential to reject real events that do not move at constant speeds. Approaches that focus on the order of cell activity within each event \cite{LeeMemorySequentialExperience2002, GuptaHippocampalReplayNot2010} allow for a relaxation of that linear assumption, but replace it with a loss of statistical power due to either ignoring all but the first spike from each cell or using an arbitrarily smoothed version of the spike train. These approaches also do not provide information about the dynamics of the underlying spatial representation. Moreover, these approaches exclude events that  have stationary representations of a single location \cite{JaiDistincthippocampalcorticalmemory2017, FarooqEmergencepreconfiguredplastic2019}.

Recognizing the problems with the linear fit, several studies have moved away from the constant velocity assumption, using an estimated represented location at each time bin and connecting each with a line. For example, using this approach, Pfeiffer and Foster \cite{PfeifferAutoassociativedynamicsgeneration2015} found that awake replays can alternate between representing a single location and sequential spatial trajectories. On the other hand, Stella et al. \cite{StellaHippocampalReactivationRandom2019} reported that replays during sleep are spatially continuous and follow Brownian diffusion dynamics. Both methods still used large time bins and neither took into account the uncertainty of the decoded estimates, however, making it hard to identify the source of the different conclusions.

An ideal approach to identifying and quantifying the dynamics of generative representations would circumvent these problems. It would use all of the available spiking data to yield the most accurate decoded positions. It would be applicable to either thresholded events or to all of the data to permit an unbiased assessment of the prevalence and nature of generative activity. It would use very small temporal bins (1 or 2 ms) to allow for very rapid representational movement and provide information about the certainty of the decoded estimates. It would be able to capture different types of movement dynamics, ranging from stationary to rapidly evolving to disorganized, and provide a statistical assessment of confidence for each category of dynamic. Finally, where assumptions are made, it would provide well defined parameters whose values can be explored systematically to understand their influence on the results.

We therefore developed a state space model of generative activity that achieves all of those goals. State space models are a well-understood, well-known statistical solution to the problems described above. By mathematically modeling the relationship between the data and latent dynamics, state space models make the assumptions of the model explicit and interpretable. Our model goes beyond previous approaches \cite{MaboudiUncoveringtemporalstructure2018, DengRapidclassificationhippocampal2016} by characterizing represented trajectories as a mixture of three underlying patterns of movement dynamics: stationary trajectories, continuous trajectories that can progress at many times the typical speed of the animal, and spatially fragmented trajectories. We show how this model can take advantage of clusterless decoding---which relates multiunit spike waveform features to position without spike sorting---giving us more information about the population spiking activity. We apply this model to data from 10 rats, focusing on SWR events to permit a direct comparison to previous work.

We find that the large majority of SWRs contain spatially coherent content; that is, the replay trajectories that are spatially concentrated at each moment in time and have no large jumps. Surprisingly, while the expected high speed, sequential replay events were identified, the most common category of events expressed representations that moved at much slower speeds consistent with real-world experiences, suggesting that high speed replay is not the most common manifestation of generative activity. These findings illustrate the power of state space models and provide a new understanding of the nature of hippocampal replay.

@@

Next, we specify movement dynamic models for how latent position---the "mental" position of the animal represented by the cells---evolves over time for a given dynamic (Figure \ref{1}C). We do this by defining a state transition matrix that defines how the latent position can move from the previous time step. Previous findings suggest that replay may exhibit at least three distinct types of movement dynamics: stationary \cite{JaiDistincthippocampalcorticalmemory2017, FarooqEmergencepreconfiguredplastic2019}, continuous \cite{DavidsonHippocampalReplayExtended2009}, and fragmented, which could correspond to both extended spatially incoherent representations and representations that jump from one position to another in a short time period \cite{PfeifferAutoassociativedynamicsgeneration2015}. We therefore define state transition models to capture each of these movement dynamics.

@@

Importantly, this approach allows the model to to capture a wide range of movement speeds for the latent position. The model is defined in terms of a mixture of movement dynamics, as summarized by the probability of each movement dynamic, and which dynamic or dynamics are dominant at a given moment is related to the movement of the underlying position representation. To demonstrate this, we applied the model to 10,000 simulated trajectories, each trajectory proceeding at a constant speed (Figure \ref{1}F) from 1 cm/s to 10,000 m/s. From this, we can see that not only are there regions of speed that correspond to our three movement dynamics being highly probable (where we define highly probable to be greater than or equal to 0.8 probability), there are intermediate speeds where two of the dynamics exhibit relatively high probability; and where the sum of two of the dynamics' probabilities exceeds 0.8. In this manuscript, we will refer to these as mixture dynamics. For example, when the stationary dynamic has a probability of 0.6 and the continuous has a probability of 0.4, we call this a stationary-continuous mixture (light blue, Figure \ref{1}G) and this indicates that the trajectory is moving slowly. Correspondingly, if the continuous dynamic has a probability of 0.5 and the fragmented dynamic has a probability of 0.4, then we would call this a fragmented-continuous-mixture and this indicates the trajectory is moving very quickly, but not as quickly as the fragmented dynamic dictates. By using this classification scheme, we can characterize the speed or set of speeds in a replay by its movement dynamics. To summarize, there are three movement dynamics, each of which is assigned a probability over time. We further classify the probability of each movement dynamic as being part of one of five speed categories: stationary, stationary-continuous-mixtures, continuous, fragmented-continuous mixtures, and fragmented.

@@

\begin{figure*}%[tbhp]
\centering
\includegraphics[width=0.80\linewidth]{figures/Figure2/Figure2_v5}
\caption{The model can decode hippocampal replay trajectories using either sorted and clusterless spikes from the same SWR event. \textbf{(A)} Decoding using sorted spikes. The top panel shows 31 cells on a W-track ordered according to linearized position by their place field peaks. The middle panel shows the probability of each dynamic over time as in Figure \ref{1}F. Shaded regions correspond to the speed classifications as in Figure \ref{1}G. The bottom panel shows the estimated probability of latent position over the course of the SWR as it travels down the center arm toward the right arm. L, R, C correspond to the position of the left, right and center reward wells respectively. The animal's actual position is indicated by the the magenta dashed line. Underneath is the maximum of the 1D decoded position (the most probable position) projected back onto the 2D track for the sorted decoding. Color indicates time. The animal's actual position is denoted by the pink dot. Light grey lines show the animal's 2D position over the entire recording session. \textbf{(B)} Decoding using clusterless spikes. The top panel shows multiunit spiking activity from each tetrode. Other panels have the same convention as (A). Underneath is the maximum of the 1D decoded position (the most probable position) projected back into 2D using the clusterless decoding.
}
\label{2}
\end{figure*}
\subsection*{Identification of continuous replays with sorted and clusterless spikes}

\begin{figure*}%[tbhp]
\centering
\includegraphics[width=0.80\linewidth]{figures/Figure3/Figure3_v5}
\caption{A-F. Six examples of SWRs with non-constant speed trajectories. Figure conventions are the same as in Figure \ref{2}. Filtered SWR ripple (150-250 Hz) trace from the tetrode with the maximum amplitude displayed above each example. \textbf{(A)} A SWR where the decoded position starts moving down the center arm away from the animal's position at the center well, slows down, and returns back. \textbf{(B)} A SWR where the decoded position persistently stays at the choice point (open circle) while the animal remains at the left well. \textbf{(C)} A SWR where the decoded position begins with stationary representation of the left well, then jumps to the middle of the right arm and proceeds up the right arm to the right well. \textbf{(D)} A SWR where the decoded position begins with stationary representation of the left well, jumps to the center arm, proceeds away from the center well, jumps to the right arm, proceeds back toward the center well, and then becomes fragmented. \textbf{(E)} A SWR where the decoded position begins in the left arm and persists at the end of the center arm. \textbf{(F)} A SWR where the decoded position starts in the left arm toward the choice point, jumps to the right arm and proceeds back toward the choice point. \textbf{(G)} Classification of SWRs from multiple animals and datasets. Each dot represents the percentage of SWRs for each day for that animal. A SWR is included in the numerator of the percentage if at any time it includes the classifications listed below the column. The denominator is listed in the x-axis label.
}
\label{3}
\end{figure*}

@@

\subsection*{Replays with changing speed}

After testing our model on a single SWR event, we applied our clusterless decoding algorithm to hippocampal recordings from 10 animals performing the W-track spatial alternation task (\# tetrodes range: [10, 24], brain areas = [CA1, CA2, CA3]; some data previously used in \cite{KarlssonAwakereplayremote2009, KayConstantSubsecondCycling2020, CarrTransientSlowGamma2012}, position linearized to 1D). One major goal of our work was to assess the overall prevalence of spatial content across SWRs, so we detected SWRs using a permissive threshold (see Methods) to ensure that we included both the rare large events as well as the much more common small events.

@@

Our model is specified in the context of a latent position associated with different movement dynamics. This specification allows us to not only to classify events in terms of their dynamics, but also to quantify the model's certainty in each position estimate at each moment in time given the model parameters. To do so we can compute the cumulative spatial extent of the 95\% highest posterior density (HPD) region of the posterior linear position. Larger values of the cumulative HPD extent indicate the model is less certain about position, because the most probable decoded positions are distributed over more of the track at a given time point. In contrast, lower values indicate that the model is more certain about the estimate of position because the extent of the HPD is more concentrated and covers less of the track. Thus, the HPD provides a complementary measure of spatial coherence, and evaluating it allows us to verify that the events we defined as spatially coherent correspond to events where there is most often high certainty around the position estimates.

Although this is certainly influenced by the dynamics imposed on the model via the movement state transition matrix, because the model is primarily driven by the data, it does not necessarily have to be the case that the two coincide. To show this, as before, we resampled position with replacement for two recording sessions, shuffling the relationship between spiking and spatial information and then fitting the encoding model. The shuffled decodes had much larger spatial coverage than the real data (XX cm vs XX cm, p=2.2e-16, one-sided Mann-Whitney-U, \ref{fig:Figure4-Figure supplement 1}).

We found that spatially coherent events did indeed have lower cumulative HPD sizes than events with fragmented dynamics. Figure \ref{4}A and B show two example SWRs that were classified as having stationary and continuous dynamics, respectively. The bulk of the HPD values at each time step in these SWRs is concentrated in a small portion of the track and the cumulative HPD size is low throughout the SWRs. In contrast, Figure \ref{4}C shows a SWR where the dynamics are fragmented and correspondingly, the HPD is much more spatially diffuse and the cumulative HPD size is much higher. The HPD also provides insights into the moment-by-moment structure of each event, which can change over the time course of a SWR. An example of this change is shown in Figure \ref{4}D, where the cumulative HPD size is small for most of the SWR until the end, at which point the uncertainty of position becomes much higher, reflecting a transition from a spatially coherent to a spatially incoherent representation. Overall, when we examined the average cumulative HPD size for each SWR, grouped by dynamic, we found a clear bimodal split between spatially coherent dynamics and spatially incoherent dynamics (Figure \ref{4}E,). For the spatially coherent dynamics, the cumulative HPD was much lower than the spatially incoherent dynamics (median 24 cm, spatially coherent vs. median 238 cm, spatially incoherent, p=2.2e-16, one-sided Mann-Whitney-U). We note here that while the size of the HPD will be influenced by the dynamic estimated from the data, it remains the case that a continuous or stationary dynamic can correspond to a large HPD region (see outliers in Fig. 4E), indicating less certainty for those specific events.

\begin{figure*}%[tb2]
\centering
\includegraphics[width=0.80\linewidth]{figures/Figure4/Figure4_v6}
\caption{Validation of classification using the 95\% Highest Posterior Density. \textbf{(A-D)}  Examples of the 95 \% Highest Posterior Density. In each column: top panel: Probability of dynamic over time. Shading and labels indicate dynamic categories. Middle panel: Posterior distribution of estimated linear position over time. Red triangle indicates animal's position. Bottom panel: Posterior distribution plot, where black designated the position bins that are included in the 95\% highest posterior density (HPD) region. 4th panel: Cumulative HPD size. \textbf{(E)} Average Cumulative 95\% HPD size for each dynamic category. Spatially coherent events (stationary, stationary-continuous, and continuous) each had much smaller median HPDs than spatially incoherent events (fragmented-continuous, fragmented, and unclassified), p < XX, ranksum tests.}
\label{4}
\end{figure*}

\subsection*{Many replay trajectories have slower movement speeds}
\begin{figure*}%[tbhp]
\centering
\includegraphics[width=0.80\linewidth]{figures/Figure5/Figure5_v4}
\caption{
Prevalence of classifications. \textbf{(A)} UpSet plot \cite{LexUpSetVisualizationIntersecting2014}---which is similar to a Venn diagram with more than 3 sets---of the most common sets of classifications within each SWR. Each row represents a classification and each column represents a set of classifications, where filled-in black dots with an edge between the dots indicates that multiple classifications are present in the SWR (at temporally distinct times). The sets are ordered by how often they occur as indicated by the bar plot above each category. The total percentage of each classification is indicated by the rotated bar plot on the left. \textbf{(B)} Duration of each dynamic category within a SWR. The box shows the interquartile range (25-75\%) of the data and the whiskers are 1.5 times the interquartile range. Points outside of this are labeled as diamonds. \textbf{(C)} Average distance between the latent position and the animal's actual position for each classification within the SWR. \textbf{(D)} Average speed of the classification within the SWR (excluding classifications < 20 ms).\textbf{(E)} Kernel density estimate of the position of stationary trajectories on the W-track at least 30 cm away from the animal's position. The shaded region represents the density estimate while the black ticks represent the observed non-local stationary positions. \textbf{(F)} Average tetrode multiunit spike rates for each dynamic category within each SWR (excluding classifications < 20 ms).
}
\label{5}
\end{figure*}
